openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-10-01 paper/policar2020-opentsne.tex
More edits to methods section

Optimization is performed with batch gradient descent using the delta-bar-delta
update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
iterations consisting of two phases: in the first \textit{early exaggeration}
phase, the attractive forces between data points is increased by some factor
$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the embedding to find
their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
which reverts the attractive forces to their original values and produces the
final embedding.

@@

\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE
gradients requires $\mathcal{O}(N^2)$ operations, which makes its application
impractical to any reasonably-sized data set and beckons for the development of
efficient approximation schemes. Van der Maaten observed that the t-SNE
gradient may be cast as an N-body problem where data points represent
particles which attract and repel each
other~\cite{van2014accelerating}. The gradient from Eqn. (\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
\right )^{-1}$. This equation can then be viewed as a particule simulation, where 
the two terms represent the attractive and repulsive
forces between individual particles. Each term lends itself to efficient
approximations, which enable us to greatly reduce the time complexity of t-SNE.

@@

Van der Maaten observed that evaluating the attractive forces between all pairs
of data points is excessive, and that considering only a handful of nearest
neighbors at each point is sufficient to obtain a good approximation~\cite{van2014accelerating}.
Because distances are transformed to similarities using a Gaussian kernel, and
the bandwidth of each kernel is selected such that only a predefined number of
neighbors fall within the main probability mass of the bell curve through the perplexity parameter, the remaining
data points fall into the exponentially decaying tails of each Gaussian distribution.
These data points are then assigned near-zero probabilities and do not contribute
to the overall attractive forces of data points.
Therefore it is sufficient to calculate the attractive forces for only a
small number of nearest neighbors instead of all $N$ points. By
utilizing tree-based nearest-neighbor search methods, the time complexity is
thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
realized that, qualitatively, embeddings are visually
indistinguishable when using only \textit{approximate} nearest neighbors,
further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each
point indiscriminately exerts a repulsive force on all other points. Van der
Maaten proposed an approach based on N-body simulations and used a
space-partitioning Barnes-Hut tree approach to approximate the interaction
between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the approach splits
the space into quadrants and entire regions may be summarized by simple statistics.
If a query point is far away from a given quadrant, the repulsive forces exerted
by all the points in that quadrant onto the query point are summarized by its center of mass. This
reduces the time complexity from $\mathcal{O}(N^2)$ to
$\mathcal{O}(N \log N)$.

@@

t-SNE is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through the use of optimization
techniques~\cite{policar2019embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place while
new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}).
Plugging these
The gradients of
$\mathbf{w}_j$ with respect to the loss (Eqn.~\ref{eq:kl_divergence}) are:
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}.
\label{eq:gradient}
\end{equation}

As in standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
\cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$
and $\mathcal{O}(\max \{ N, M \})$, respectively. In the FIt-SNE approximation
scheme, we additionally exploit the fact that the reference embedding remains
fixed throughout the optimization of newly added points, and precompute the
interpolation grid. This further reduces the runtime complexity from
$\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.

@@

In standard t-SNE, distances are converted to similarities through the use of
Gaussian kernels of varying bandwidths. The bandwidths are indirectly
determined by the user-specified perplexity parameter so that a fixed number of
nearest data points will have non-zero similarity. One common trick for
uncovering the global relations between clusters is to increase perplexity so
that more long-range interactions are preserved in the final embedding.
However, one unfortunate side effect of increasing perplexity is that smaller
clusters get absorbed into larger ones.

Kobak \& Berens suggest that replacing the Gaussian kernel
with a mixture of Gaussians may provide better insight into both the local and
global structure~\cite{kobak2019art}. Therefore, the similarities between data points in the input
space are computed with

@@

Standard t-SNE reveals the clustering structure at a single level of
resolution. Different perplexity parameter values can be used to identify
global cluster relationships or small, well-isolated groups. Unfortunately,
varying perplexity values can be time-consuming as this involves recomputing
the KNN graph which is often the most expensive part of the t-SNE algorithm.
Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying
the degree of freedom in the t-distribution can be used to explore the clustering structure at different
levels of resolution.

Standard t-SNE models similarities between data points in the embedding space
using a t-distribution with a single degree of freedom

\begin{equation}
q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha} = \frac{1}{\left( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right)^\alpha }.
\end{equation}
In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from
Eqn.~(\ref{eq:cauchy_kernel}). The gradient of the loss function then becomes
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) w_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
\end{align}
which can, again, be cast as attractive and repulsive forces between particles
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} w_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
&- 4 \sum_{j \neq i} w_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
\end{align}

@@

B{\"o}hm \textit{et al.} showed that the exaggeration factor
$\rho$ can be used to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
By incorporating exaggeration into later phases of optimization, t-SNE is able
introduce more white-space between clusters and better reflect the global
relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and $\rho=30$
produces embeddings visually similar to UMAP and ForceAtlas2, respectively.

While it is difficult to claim one is better than the other, different settings
of $\rho$ may uncover different properties and clustering structures. For
example, in single-cell data, standard t-SNE with $\rho=1$ often uncovers
distinct groups of cell-types but obscures developmental paths. This problem is
exacerbated when dealing with large numbers of data points. On the other hand,
ForceAtlas2 has often been used to uncover trajectories and transitions between
cell types, but larger clusters often absorb small, distinct groups of cells.
Therefore one may run t-SNE with two different exaggeration settings, one to
highlight the developmental transitions between cell types and one to identify
clear populations of cells.

@@

t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout algorithms,
which operate on the $k$-nearest neighbor graph. Each method constructs the graph
and specifies the attractive and repulsive forces in its own way, but ultimately,
point positions are found by balancing the attractive and repulsive forces between data points.

The final positioning of data points using this class of algorithms is largely
dependent on the embedding initialization. UMAP found early success as it was
able to produce visualizations that better captured the global organization of
clusters. This was due to its initialization scheme, which initialized point
positions using Laplacian eigenmaps. On the other hand, most implementations of
t-SNE performed random initialization, which resulted in poor global coherence.
However, Kobak \& Linderman recently showed that using the same initialization
for both methods leads to visualizations, which exhibit similarly good global
coherency. Conversely, when initialized randomly, both methods produce
visualizations where clusters are arbitrarily positioned in the embedding space.

@@

openTSNE is distributed under the BSD-3-Clause License and is publicly
available as an open-source package at
\url{https://github.com/pavlin-policar/openTSNE}. openTSNE is also available on
PyPI and conda-forge. The data sets used and scripts used in this study are
included in accompanying notebooks, publicly available at
\url{https://github.com/pavlin-policar/opentsne-paper}.

>>>>

Optimization is performed with batch gradient descent using the delta-bar-delta
update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
iterations consisting of two phases: in the first \textit{early exaggeration}
phase, the attractive forces between data points are increased by some factor
$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and find
their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
which reverts the attractive forces to their original values and produces the
final embedding.

@@

\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE
gradients requires $\mathcal{O}(N^2)$ operations, which makes its application
impractical to any reasonably-sized data set and beckons for the development of
efficient approximation schemes. Van der Maaten observed that the t-SNE
gradient may be cast as an N-body problem where data points represent
particles which attract and repel each
other~\cite{van2014accelerating}. The gradient from Eqn. (\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
\right )^{-1}$. This equation can then be viewed as a particle simulation, where 
the two terms represent the attractive and repulsive
forces between individual particles. Each term lends itself to efficient
approximations, which enable us to greatly reduce the time complexity of t-SNE.

@@

Van der Maaten observed that evaluating the attractive forces between all pairs
of data points is excessive, and that considering only a handful of nearest
neighbors at each point is sufficient to obtain a good approximation~\cite{van2014accelerating}.
Because distances are transformed to similarities using a Gaussian kernel, and
the bandwidth of each kernel is selected such that only a predefined number of
neighbors fall within the main probability mass of the bell curve through the perplexity parameter, the remaining
data points fall into the exponentially decaying tails of each Gaussian distribution.
These data points are assigned near-zero probabilities and do not contribute
to the overall attractive forces of data points.
Therefore it is sufficient to calculate the attractive forces for only a
small number of nearest neighbors instead of all $N$ points. By
utilizing tree-based nearest-neighbor search methods, the time complexity is
thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
realized that, qualitatively, embeddings are visually
indistinguishable when using only \textit{approximate} nearest neighbors,
further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each
point indiscriminately exerts a repulsive force on all other points. Van der
Maaten proposed an approach based on N-body simulations and used a
space-partitioning Barnes-Hut tree approach to approximate the interaction
between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the approach splits
the space into quadrants and entire regions may be summarized by simple statistics.
If a query point is far away from a given quadrant, the repulsive forces exerted
by all the points in that quadrant onto the query point are summarized by a single point. This
reduces the time complexity from $\mathcal{O}(N^2)$ to
$\mathcal{O}(N \log N)$.

@@

t-SNE is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through the use of optimization
techniques~\cite{policar2019embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place while
new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}).
Plugging these terms into Eqn.~(\ref{eq:kl_divergence}), we obtain the following
gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}.
\label{eq:gradient}
\end{equation}

Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
\cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$
and $\mathcal{O}(\max \{ N, M \})$, respectively. In the FIt-SNE approximation
scheme, we additionally exploit the fact that the reference embedding remains
fixed throughout the optimization of newly added points, and precompute the
interpolation grid. This further reduces the runtime complexity from
$\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.

@@

In standard t-SNE, distances are converted to similarities through the use of
Gaussian kernels of varying bandwidths. The bandwidths are indirectly
determined by the user-specified perplexity parameter so that a fixed number of
nearest data points will be assigned non-zero values. One common trick for
uncovering the global relations between clusters is to increase perplexity so
that more long-range interactions are preserved in the final embedding.
However, one unfortunate side effect of increasing perplexity is that smaller
clusters get absorbed into larger ones.

Kobak \& Berens suggest that replacing the Gaussian kernel
with a mixture of Gaussians may provide better insight into both the local and
global structure~\cite{kobak2019art}. For instance, the similarities between data points in the input
space may instead be computed with

@@

Standard t-SNE reveals the clustering structure at a single level of
resolution. Different perplexity parameter values can be used to identify
global cluster relationships or small, well-isolated groups. Unfortunately,
varying perplexity values can be time-consuming as this involves recomputing
the $k$-nearest neighbor graph which is often the most expensive part of the t-SNE algorithm.
Alternatively, Kobak \textit{et al.} suggest that varying
the degree of freedom in the t-distribution can be used to explore the clustering structure at different
levels of resolution~\cite{kobak2019heavy}.

Standard t-SNE models similarities between data points in the embedding space
using a t-distribution with a single degree of freedom, but this can be generalized
to allow for any parameter value

\begin{equation}
q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha} = \frac{1}{\left( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right)^\alpha }.
\end{equation}
In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from
Eqn.~(\ref{eq:cauchy_kernel}). The gradient of the loss function then becomes
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) w_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
\end{align}
which can, again, be cast as the interplay between the attractive and repulsive forces between particles
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} w_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
&- 4 \sum_{j \neq i} w_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
\end{align}

@@

B{\"o}hm \textit{et al.} showed that the exaggeration factor
$\rho$ can be used to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
By incorporating exaggeration into later phases of optimization, t-SNE is able
introduce more white-space between clusters which better reflects the global
relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and $\rho=30$
produces embeddings visually similar to UMAP and ForceAtlas2, respectively.

While it is difficult to claim one is better than the other, different settings
of $\rho$ may uncover different properties of the data manifold. For
example, in single-cell data, standard t-SNE often uncovers
distinct, well-separated groups of cell-types but obscures transitional paths between cell-states. This problem is
exacerbated when dealing with large numbers of data points. On the other hand,
ForceAtlas2 has been used successfuly to uncover trajectories and transitions between
cell types. Unfortunately, when using ForceAtlas2, large clusters often absorb small, distinct groups of cells.
One might run t-SNE with various parameter settings to avoid switching between
different algorithms and obtain similar visualizations.
We might generate an embedding using higher levels of exaggeration to highlight
the developmental transitions between cell types and
one with low levels of exaggeration to identify clear populations of cells.

@@

t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout algorithms,
which operate on the $k$-nearest neighbor graph. Each method constructs the graph differently
and specifies the attractive and repulsive forces in its own way, but ultimately,
point positions are found by balancing the attractive and repulsive forces between data points.

The final positions of data points in this class of algorithms is largely
dependent on the embedding initialization. UMAP found early success as it was
able to produce visualizations that better captured the global organization of
clusters. This was due to its initialization scheme, which initialized point
positions using Laplacian eigenmaps. On the other hand, most implementations of
t-SNE performed random initialization, which resulted in poor global coherence.
However, Kobak \& Linderman recently showed that using the same initialization
for both methods leads to visualizations, which exhibit similarly good global
coherency. Conversely, when initialized randomly, both methods produce
visualizations where clusters are arbitrarily positioned in the embedding space.

@@

\opentsne\ is distributed under the BSD-3-Clause License and is publicly
available as an open-source package at
\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available on
\texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used in this study are
included in accompanying notebooks, publicly available at
\url{https://github.com/pavlin-policar/opentsne-paper}.
